---
title: "Interactive Map Making"
author: "Kanika Khanna"
date: "`r Sys.Date()`"
output: html_document
---
When working with geospatial data and analyses, we often find that illusstrating findings and patterns with maps can be incredibly useful. R packages like tidyverse and ggplot2 among others are helpful to create clear, clean, static maps, perfectly suitable for publications. But what if you're interested in embedding a map into a website or blog post? How can we make our geospatial data more interactive?

Using leaflet, we can create simple and effective interactive maps of our data in R, without relying on additional programs, services, or software to do so. Today's workshop will cover some of the basics of this process. We will use some data that I have collected and cleaned for these exercises. They will be available for you to use to follow along today.

First, let's load the libraries we are using. If you don't have them installed, please do so!

```{r}

#load libraries
library(acs)
library(tigris)
library(leaflet)
library(mapview)
library(stringr)
library(dplyr)
options(tigris_use_cache = TRUE) #I tend to cache the tigris data here so that my computer is not constantly fetching the data; it slows it down otherwise. You do you.

```

Next, let's pull up leaflet tiles. Because the data I am using today focuses on New York City, I've put the setView to zoom in on it -- the order here is x, y, longitude then latitude, so be mindful when you input things in like this!
```{r}
leaflet() %>%
  addTiles() %>%
  setView(-74.00, 40.71, zoom = 12) %>%
  addProviderTiles("CartoDB.Positron")
```

```{r}
#The tigris package makes it easy to grab Census shapefiles at various levels, from tracts and zipcodes all the way up to national borders. First we look up the appropriate codes.
lookup_code("New York", "New York")
lookup_code("New York", "Kings")
lookup_code("New York", "Queens")
lookup_code("New York", "Bronx")
lookup_code("New York", "Richmond")
```

```{r}
# The summary below shows the data associated with each polygon (e.g., the GEOID and  NAME). I can use this to build the Census tract GEOIDs that I will later join my data by, as it is in the same form.
nyc_tracts <- tracts(state = '36', county = c('061','047','081','005','085'))

```
```{r}
# Here's what that looks like: a real map structure! Yes this is very exciting to me.
plot(nyc_tracts)
```

```{r}
#bringing in my own data
housing_data <- read.csv("TTCDATA.csv")

class(housing_data$pct.pov)

#mutating the percentage variables
housing_data_clean <- housing_data %>% 
mutate(pct.asian=pct.asian*100, 
       pct.black=pct.black*100,
       pct.latinx=pct.latinx*100,
       pct.white=pct.white*100,
       pct.pov=(as.numeric(housing_data$pct.pov)*100),
       pct.min= pct.black+pct.latinx+pct.asian)

summary(housing_data_clean)
class(housing_data_clean$GEOID)
housing_data_clean$GEOID <- as.character(housing_data_clean$GEOID)

```
```{r}
#merging the data!
mydatamerged <- right_join(nyc_tracts, housing_data_clean, "GEOID", "GEOID")
                        
#making sure all is in order...
summary(mydatamerged)
#yay it worked!
```

```{r}
#renaming my dataframe
nyc_housingtracts <- mydatamerged
```

```{r}
#time to start mapping!

#mapping poverty
mypopup_pov <- paste0("GEOID: ", nyc_housingtracts$GEOID, "<br>", "Poverty Rate: ", round(nyc_housingtracts$pct.pov,0))

#colors and bin setting
mypovbin <-c(0,10,20,30,40,50,60,70,80,90)
mypal_pov <- colorBin(
  palette = "OrRd",
  domain = nyc_housingtracts$pct.pov,
  bins=mypovbin
)

#using leaflet to generate my map
povmap<-leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_housingtracts, 
              fillColor = ~mypal_pov(pct.pov), 
              color = "#b2aeae", #drawing the lines for the Census tract boundaries -- have to use hex colors for this
              fillOpacity = 0.7, 
              weight = 1, 
              smoothFactor = 0.2,
              popup = mypopup_pov) %>%
  addLegend(pal = mypal_pov, #drawing the legend
            values = nyc_housingtracts$pct.pov, 
            position = "bottomright", 
            title = "Poverty Rate",
            labFormat = labelFormat(suffix = "%"))


povmap


```


```{r}
#mapping racial and ethnic groups! I'll start with white
mypopup_white <- paste0("GEOID: ", nyc_housingtracts$GEOID, "<br>", "Percent White ", round(nyc_housingtracts$pct.white,0))

#colors and bins
mybins=c(0,10,20,30,40,50,60,70,80, 90, 100)
mypal_white <- colorBin(
  palette = "PuRd",
  domain = nyc_housingtracts$pct.white,
  bins=mybins
)

leaflet
whitemap<-leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_housingtracts, 
             fillColor = ~mypal_white(pct.white), 
             color = "#b2aeae", # you need to use hex colors
             fillOpacity = 0.7, 
             weight = 1, 
             smoothFactor = 0.2,
             popup = mypopup_white) %>%
  addLegend(pal = mypal_white, 
           values = nyc_housingtracts$pct.white, 
           position = "bottomright", 
           title = "Percent White",
           labFormat = labelFormat(suffix = " %"))

whitemap
```
```{r}
#mapping black population
mypopup_black <- paste0("GEOID: ", nyc_housingtracts$GEOID, "<br>", "Percent Black ", round(nyc_housingtracts$pct.black,0))

#colors and bins
mybins=c(0,10,20,30,40,50,60,70,80, 90, 100)
mypal_black <- colorBin(
  palette = "YlOrBr",
  domain = nyc_housingtracts$pct.black,
  bins=mybins
)

leaflet
blackpopmap<-leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_housingtracts, 
             fillColor = ~mypal_black(pct.black), 
             color = "#b2aeae", # you need to use hex colors
             fillOpacity = 0.7, 
             weight = 1, 
             smoothFactor = 0.2,
             popup = mypopup_black) %>%
  addLegend(pal = mypal_black, 
           values = nyc_housingtracts$pct.black, 
           position = "bottomright", 
           title = "Percent Black",
           labFormat = labelFormat(suffix = " %"))
blackpopmap
```
```{r}
#mapping Asian
mypopup_asian <- paste0("GEOID: ", nyc_housingtracts$GEOID, "<br>", "Percent Asian ", round(nyc_housingtracts$pct.asian,0))

#colors
mybins=c(0,10,20,30,40,50,60,70,80, 90, 100)
mypal_asian <- colorBin(
  palette = "BuGn",
  domain = nyc_housingtracts$pct.asian,
  bins=mybins
)

leaflet
asianmap<-leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_housingtracts, 
             fillColor = ~mypal_asian(pct.asian), 
             color = "#b2aeae", # you need to use hex colors
             fillOpacity = 0.7, 
             weight = 1, 
             smoothFactor = 0.2,
             popup = mypopup_asian) %>%
  addLegend(pal = mypal_asian, 
           values = nyc_housingtracts$pct.asian, 
           position = "bottomright", 
           title = "Percent Asian",
           labFormat = labelFormat(suffix = " %"))

asianmap
```
```{r}
#mapping Latinx
mypopup_latinx <- paste0("GEOID: ", nyc_housingtracts$GEOID, "<br>", "Percent Latinx ", round(nyc_housingtracts$pct.latinx,0))

#colors
mybins=c(0,10,20,30,40,50,60,70,80, 90, 100)
mypal_latinx <- colorBin(
  palette = "BuPu",
  domain = nyc_housingtracts$Tpct.latinx,
  bins=mybins
)

leaflet
latinxmap<-leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = nyc_housingtracts, 
             fillColor = ~mypal_latinx(pct.latinx), 
             color = "#b2aeae", # you need to use hex colors
             fillOpacity = 0.7, 
             weight = 1, 
             smoothFactor = 0.2,
             popup = mypopup_latinx) %>%
  addLegend(pal = mypal_latinx, 
           values = nyc_housingtracts$pct.latinx, 
           position = "bottomright", 
           title = "Percent Latinx",
           labFormat = labelFormat(suffix = " %"))

latinxmap
```

Now, these are examples of chloropleth maps draw in leaflet. What if you have other data you want to map in a different way?

Here I'm going to use some data on Confederate monuments in the US. The geographic variable here is coordinate points.

```{r}
#loading the file
monuments <-read.csv("8-29 Monuments Revgeocoded.csv")
```

We have to load new tiles!

```{r}
monuments %>% 
leaflet() %>%
addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") %>%
addProviderTiles(providers$Stamen.TonerLite, group = "Toner Lite") %>%
addLayersControl(baseGroups = c("Toner Lite", "World Imagery"))
```

```{r}
monuments %>% 
leaflet() %>%
addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") %>%
addProviderTiles(providers$Stamen.TonerLite, group = "Toner Lite") %>%
addLayersControl(baseGroups = c("Toner Lite", "World Imagery")) %>%
  addCircleMarkers(lng = ~longitude, lat = ~latitude, 
             popup = ~Monument_Name)  
```








